Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-05-28 ebayes_rnaseq_heterosis.tex
need to update results

We present a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.

@@

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values method of \cite{robinson2010scaling}. We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$. Using the gene-specific parameter estimates, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g. $\hat{\eta}_\delta = \sum_{g=1}^G \hat{\delta}_g/G$ and $2\hat{\sigma}_\delta^2 = \sum_{g=1}^G (\hat{\delta}_g - \hat{\eta}_\delta)^2/(G-1)$.

@@

We used our method to analyze the maize data set, generated by \cite{paschold2012complementation}, on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). The data are available in the Supplemental Materials. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

@@

The supplementary materials include the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}), and the {\tt Stan} model (\verb@sg_model_ind.txt@).  
%the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.

>>>>

We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.

@@

To obtain normalization factors $\hat{\gamma}$, we use the weighted trimmed mean of $M$ values (TMM) method of \cite{robinson2010scaling}. We then use the methods of \cite{mccarthy2012differential}, to obtain genewise dispersion estimates, $\hat{\phi}_g$, and their generalized linear model methods to obtain estimates for the remaining gene-specific parameters, i.e. $\hat{\phi}_g$, $\hat{\alpha}_g$, and $\hat{\delta}_g$. Using the gene-specific parameter estimates, we estimate hyperparameters for the location and scale parameters in the hierarchical model using a central method of moments approach, e.g. $\hat{\eta}_\delta = \sum_{g=1}^G \hat{\delta}_g/G$ and $2\hat{\sigma}_\delta^2 = \sum_{g=1}^G (\hat{\delta}_g - \hat{\eta}_\delta)^2/(G-1)$.

@@

We used our method to analyze a maize data set \citep{paschold2012complementation} on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). The data are available in the Supplemental Materials. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

@@

The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the {\tt Stan} model ({\tt model.stan}).  
%the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.
